### 2025-08-27 sections/1-introduction.tex
orphans, whitespace, lfd plots

Hierarchical Navigable Small World networks~\cite{malkov2016hnsw} rely on navigable small world (NSW) networks and skip lists. InVerted File indexing (IVF)~\cite{faissivf, sacks1987multikey, kent1990signature} partitions data into high-dimensional Voronoi cells, and searches exhaustively only within cells near the query. ANNOY~\cite{annoy} uses random projection and tree building.
At each intermediate node of the tree, a random hyperplane, defined by two sampled points, splits the space.  Multiple such trees form a forest; increasing the number of trees improves recall but increases search time.

@@

In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search)~\cite{ishaq2019clustered}, which extended the original flat clustering approach to entropy-scaling $\rho$-NN search from a flat clustering approach~\cite{yu2015entropy} to a hierarchical one.
CLAM (Clustering, Learning and Approximation with Manifolds), originally developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, refined the clustering algorithm from CHESS.

This paper introduces CAKES, a set of three entropy-scaling algorithms for $k$-NN search, implemented in Rust. We also provide a theoretical analysis of their time complexity in Sections~\ref{sec:methods:knn-search:repeated-rnn-complexity} and~\ref{sec:methods:knn-search:complexity-of-sieve-methods}.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a uniform distribution.
Given that CAKES's algorithms are intended to be used on datasets with a manifold structure, complexity analysis assuming a uniform distribution of data would be uninformative.
As a result, our analysis assumes datasets with manifold structure—characterized by low fractal dimension and metric entropy—reflecting the conditions for which CAKES is designed.

>>>>

Hierarchical Navigable Small World networks~\cite{malkov2016hnsw} rely on navigable small world (NSW) networks and skip lists. InVerted File indexing (IVF)~\cite{faissivf, sacks1987multikey, kent1990signature} partitions data into high-dimensional Voronoi cells, and searches exhaustively only within cells near the query. ANNOY~\cite{annoy} uses random projection and tree building.
At each intermediate node of the tree, a random hyperplane, defined by two sampled points, splits the space.  Increasing the number of such trees improves recall but increases search time.

@@

In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search)~\cite{ishaq2019clustered}, which extended the flat clustering approach used in~\cite{yu2015entropy} to a hierarchical one.
CLAM (Clustering, Learning and Approximation with Manifolds), originally developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, refined the clustering algorithm from CHESS.

This paper introduces CAKES, a set of three entropy-scaling algorithms for $k$-NN search, implemented in Rust. We also provide a theoretical analysis of their time complexity in Sections~\ref{sec:methods:knn-search:repeated-rnn-complexity} and~\ref{sec:methods:knn-search:complexity-of-sieve-methods}.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a uniform distribution.
Given that CAKES's algorithms are intended for datasets with a manifold structure, complexity analyses assuming a uniform distribution would be uninformative.
As a result, our analysis assumes datasets with manifold structure—characterized by low fractal dimension and metric entropy—reflecting the conditions for which CAKES is designed.

### 2025-08-27 sections/2-methods.tex
orphans, whitespace, lfd plots

CAKES derives its advantages from the manifold hypothesis~\cite{fefferman2016testing}, the notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space. We define the \emph{local fractal dimension} (LFD) to quantify this property.

For a point $q \in \textbf{X}$ and radii $r_1 > r_2$,

@@

\begin{figure}[H]
    \vskip -0.2in
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.pdf}
    \caption{
        {\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.
        ${\color{blue}\delta} = f(q, c)$ is the distance from the query to the cluster center $c$.
        ${\color{red}\delta^{+}} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
        ${\color{green}\delta^{-}} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
    }
    \label{fig:methods:deltas}
    \vskip -0.5in
\end{figure}

\begin{minipage}{.425\textwidth}
    \begin{algorithm}[H]
        \caption{tree-search($C$, $q$, $r$)}
        \label{alg:methods:rnn-search:tree-search}
        \begin{algorithmic}
            \REQUIRE $f$, a distance function
            \REQUIRE $C$, a cluster
            \REQUIRE $q$, a query
            \REQUIRE $r$, a search radius
            \IF{$\delta^+_C \leq r$}
                \STATE \textbf{return} $\{C\}$
            \ELSE
                \STATE $[L, R]$ $\Leftarrow$ \textit{children} of $C$
                \STATE \textbf{return} tree-search($L, q, r$) \\
                 $\cup$ tree-search($R, q, r$)
            \ENDIF
        \end{algorithmic}
    \end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{.475\textwidth}
    \begin{algorithm}[H]
        \caption{leaf-search($Q$, $q$, $r$)}
        \label{alg:methods:rnn-search:leaf-search}
        \begin{algorithmic}
            \REQUIRE $f$, a distance function
            \REQUIRE $Q$, a set of clusters
            \REQUIRE $q$, a query
            \REQUIRE $r$, a search radius
            \STATE $H \Leftarrow \emptyset$
            \FOR{$C \in Q$}
                \IF{$\delta^+_C \leq r$}
                    \STATE $H$ $\Leftarrow$ $H \cup C$
                \ELSE
                    \FOR{$p \in C$}
                        \IF{$f(p, q) \leq r$}
                            \STATE $H$ $\Leftarrow$ $H \cup \{p\}$
                        \ENDIF
                    \ENDFOR
                \ENDIF
            \ENDFOR
            \STATE \textbf{return} $H$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}

>>>>

CAKES derives advantages from the manifold hypothesis~\cite{fefferman2016testing}, the notion that high-dimensional data collected from constrained generating phenomena often occupy a low-dimensional manifold within their embedding space. We define the \emph{local fractal dimension} (LFD) to quantify this property. For a point $q \in \textbf{X}$ and radii $r_1 > r_2$,

@@

\begin{figure}[t]
    \vskip -0.2in
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.pdf}
    \caption{
        {\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.
        ${\color{blue}\delta} = f(q, c)$ is the distance from the query to the cluster center $c$.
        ${\color{red}\delta^{+}} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
        ${\color{green}\delta^{-}} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
    }
    \label{fig:methods:deltas}
    % \vskip -0.5in
\end{figure}
\FloatBarrier

% \begin{minipage}{.425\textwidth}
\FloatBarrier
    \begin{algorithm}[!t]
        \caption{tree-search($C$, $q$, $r$)}
        \label{alg:methods:rnn-search:tree-search}
        \begin{algorithmic}[1]
            \REQUIRE $f$, a distance function
            \REQUIRE $C$, a cluster
            \REQUIRE $q$, a query
            \REQUIRE $r$, a search radius
            \IF{$\delta^+_C \leq r$}
                \STATE \textbf{return} $\{C\}$
            \ELSE
                \STATE $[L, R]$ $\Leftarrow$ \textit{children} of $C$
                \STATE \textbf{return} tree-search($L, q, r$) \\
                 $\cup$ tree-search($R, q, r$)
            \ENDIF
        \end{algorithmic}
    \end{algorithm}
% \end{minipage}

% \begin{minipage}{.475\textwidth}
    \begin{algorithm}[!t]
        \caption{leaf-search($Q$, $q$, $r$)}
        \label{alg:methods:rnn-search:leaf-search}
        \begin{algorithmic}[1]
            \REQUIRE $f$, a distance function
            \REQUIRE $Q$, a set of clusters
            \REQUIRE $q$, a query
            \REQUIRE $r$, a search radius
            \STATE $H \Leftarrow \emptyset$
            \FOR{$C \in Q$}
                \IF{$\delta^+_C \leq r$}
                    \STATE $H$ $\Leftarrow$ $H \cup C$
                \ELSE
                    \FOR{$p \in C$}
                        \IF{$f(p, q) \leq r$}
                            \STATE $H$ $\Leftarrow$ $H \cup \{p\}$
                        \ENDIF
                    \ENDFOR
                \ENDIF
            \ENDFOR
            \STATE \textbf{return} $H$
        \end{algorithmic}
    \end{algorithm}
% \end{minipage}

### 2025-08-27 sections/3-datasets.tex
orphans, whitespace, lfd plots

% \subsection{Local Fractal Dimension}
% \label{sec:dayasets:lfd-of-datasets}

>>>>

\subsection{Local Fractal Dimension}
\label{sec:dayasets:lfd-of-datasets}

\begin{figure}[t]
    \centering
    % Row 1
    \subfloat[Fashion-MNIST\label{fig:results:fashion-mnist-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/fashion-mnist.png}}
    \hfill
    \subfloat[Glove-25\label{fig:results:glove-25-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/glove-25.png}}

    % Row 2
    \par\medskip
    \subfloat[Sift\label{fig:results:sift-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/sift.png}}
    \hfill
    \subfloat[Random dataset\label{fig:results:random-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/random.png}}

    % Row 3
    \par\medskip
    \subfloat[Silva 18S\label{fig:results:silva-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/silva-SSU-Ref.png}}
    \hfill
    \subfloat[RadioML\label{fig:results:radioml-lfd}]{
      \includegraphics[width=0.48\columnwidth]{images/lfd/radio-ml.png}}

    % Legend
    \par\smallskip
    \includegraphics[width=0.65\columnwidth]{images/lfd/legend.png}

    \caption{Local fractal dimension vs. cluster depth across six datasets. The ‘random’ dataset is generated as in Sec.~\ref{sec:datasets-and-benchmarks:random-datasets}; note the different y-axis. The x-axis is tree depth; the y-axis is LFD at that depth. We plot the 5th, 25th, 50th, 75th, and 95th percentiles, plus min/max. Each cluster is counted by its cardinality so the curves reflect dataset-level distributions.}
    \label{fig:results:lfd-plots}
  \end{figure}
